Measuring neutron star tidal deformability with Advanced LIGO: a Bayesian analysis of neutron star - black hole binary observations

### 2017-01-10 Paper/journal_submission/paper.tex
final version for resubmission

\begin{abstract}
The pioneering discovery of gravitational waves (GW) by Advanced LIGO has ushered
us into an era of observational GW astrophysics. Compact binaries remain the 
primary target sources for GW observation, of which neutron star - black hole 
(NSBH) binaries form an important subset.
% 
GWs from NSBH sources carry signatures of (a) the tidal distortion of the
neutron star by its companion black hole during inspiral, and (b) its
potential tidal disruption near merger.
%
In this paper, we present a Bayesian study of the measurability of
neutron star tidal deformability $\lambdans\propto (R/M)_\mathrm{NS}^{5}$
using observation(s) of inspiral-merger GW signals from disruptive NSBH
coalescences, taking into account the crucial effect of black hole spins.
% 
First, we find that if non-tidal templates are used to estimate source 
parameters for an NSBH signal, the bias introduced in the estimation of
non-tidal physical parameters will only be significant for loud signals with
signal-to-noise ratios greater than $\simeq30$.
% 
For similarly loud signals, we also find that we can begin to put
interesting constraints on $\lambdans$ (factor of $1-2$) with individual
observations.
% 
Next, we study how a population of realistic NSBH detections will improve our
measurement of neutron star tidal deformability. For an astrophysically likely
population of {\it disruptive} NSBH coalescences, we find that $20-35$ events
are sufficient to constrain $\lambdans$ within $\pm 25-50\%$, depending on the
neutron star equation of state. For these calculations we assume that LIGO will
detect black holes with masses within the astrophysical {\it mass-gap}. In case
the mass-gap remains {\it preserved} in NSBHs detected by LIGO, we estimate that
approximately $25\%$ additional detections will furnish comparable $\lambdans$
measurement accuracy.
% 
We find these results encouraging, and recommend that an effort to 
measure $\lambdans$ be planned for upcoming NSBH observations with the
LIGO-Virgo instruments.
\end{abstract}

@@

The EoS of matter at nuclear densities is believed to be universal among all
neutron stars. The Tolman-Oppenheimer-Volkoff equation~\cite{Tolman:1939jz,
Oppenheimer:1939ne,1934PNAS...20..169T}
would then predict that NS properties satisfy a universal relationship
between $\lambdans$ and $\mns$. As the final part of
this paper, we combine information from multiple observations of realistic NSBH
systems and perform a fully-Bayesian analysis of how our estimation of
$\lambdans$ changes as we accumulate detections. This is similar to an earlier
study~\cite{DelPozzo:13} aimed at binary neutron stars.
% % 
We restrict ourselves to a population of NSs with masses clustered very tightly
around $1.35M_\odot$ (with a negligible variance), and negligible spins. We
sample different nuclear EoSs by sampling entire populations fixing different
values for the NS tidal deformability.
% These values correspond to a hard EoS at the lower end, and a soft EoS at the upper.
For all populations, we take source locations to be uniformly distributed in
spatial volume, and source orientations to be uniform on the $2-$sphere. To
summarize, we find the following:
% 
(a) Our median estimate for $\lambdans$ starts out prior dominated, but 
converges to within $10\%$ of the true value within $10-20$ detections.
(b) Measurement uncertainties for $\lambdans$, on the other hand, depend on
$\lambdans$ itself. We find that for hard equations of state (with 
$\lambdans\geq 1000$), $10-20$ observations are sufficient to constrain
$\lambdans$ within $\pm 50\%$. For softer equations of state, the same level
of certainty would require substantially more ($25-40$) observations.
% 
(c) Further, if the astrophysical ``mass-gap''~\cite{Bailyn:1997xt,
Kalogera:1996ci,Kreidberg:2012,Littenberg:2015tpa} is real, we find that $20-50\%$
additional observations would be required to attain the same measurement
accuracy as above. And, (d) putting tighter constraints on the $\lambdans$ of a
population would require $50+$ NSBH observations, in any scenario.
% 
All of the above is possible within a few years of design
aLIGO operation~\cite{Abadie:2010cfa}.

@@

@@

To summarize, in this section we study the improvement in our measurement of 
NS deformability parameter $\lambdans$ with an increasing number of events. We
do so by simulating plausible populations of disrupting NSBH binaries (with
$\rho\geq 10$). We find that:
(i) for more deformable neutron stars (harder equation of states), the median
value of $\lambdans$ comes within $10\%$ of the true value with as 
few as $10$ events, while achieving the same accuracy for softer equations of 
state will take $15-20$ source detections; (ii) the statistical uncertainty
associated with $\lambdans$ measurement shrinks to within $\pm50\%$ with
$10-20$ events, and to within $\pm 25\%$ with $50+$ events, when source 
$\lambdans\geq 1000$; (iii) for softer equations of state, the same could take
$25-40$ and $50+$ events, respectively for the two uncertainty thresholds;
and (iv) if BHs really do observe the astrophysical mass-gap, the information
accumulation is somewhat slower than if they do not. We conclude that within
$20-30$ observations, aLIGO would begin to place very interesting bounds on 
the NS deformability, which would allow us to rule out or rank different
equations of state for neutron star matter. Our key findings are 
summarized in Fig.~\ref{fig:TT_LambdaMedian_vs_N_AllInOne} and
\ref{fig:TT_LambdaError_vs_N_L500_2000_CI90_0_AllInOne}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}\label{s1:discussion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% 
\begin{figure*}
\centering    
% \textbf{No Mass-Gap \hspace{6cm} Mass-Gap}\par\medskip
\includegraphics[width=1.025\columnwidth,trim=1cm 0 0 0]{LambdaMedianCIWidths90pc_vs_N_AllPopulation_SNRSorted}
\includegraphics[width=1.025\columnwidth,trim=0 0 1cm 0]{LambdaCIWidths90pc_vs_N_AstroPopulation_SNRSorted}
\caption{This figure is similar to Fig.~\ref{fig:TT_LambdaError_vs_N_L500_2000_CI90_0_AllInOne},
with the difference
that events in each population have been sorted according to their
signal strength (SNR). We note that information about
the tidal deformability of neutron stars comes primarily 
from the loudest $5-10$ events, whether we allow BH masses
in the mass gap (left panel) or restrict them to
$\mbh\geq 5M_\odot$ (right panel).
}
\label{fig:TT_LambdaError_vs_N_L500_2000_CI90_0_AllInOne_SNRSorted}
\end{figure*}
%

@@

We probe two astrophysical paradigms, one that allows for BH masses to lie
within the astrophysical mass-gap (paradigm A), and one that does not (paradigm
B).
% 
{\it For paradigm A}, we find the following: (i) for the softer equations of
state that result in less deformable neutron stars, $15-20$ detections bring
the measured probability distribution for $\lambdans$ entirely within the prior,
which ensures that the median $\lambdans$ tracks the true value to within $10\%$.
(ii) For NSBH populations with more deformable NSs ($\lambdans> 1000$),
the same is achievable within as few as $10$ (or $15$ at most) realistic
observations. (iii) The statistical uncertainty associated with $\lambdans$
measurement can be restricted to be within $\pm50\%$ using $10-20$ observations
when $\lambdans> 1000$), and using $25-40$ observations for softer equations
of state. All of the above is possible within a few years of design
aLIGO operation~\cite{Abadie:2010cfa}, if astrophysical BHs are allowed
masses $< 5M_\odot$ (i.e. in the mass-gap). However, further
restricting $\lambdans$ will require $50+$ NSBH observations.
% 
{\it For paradigm B}, we find the information accumulation to be somewhat slower.
While the quantitative inferences for populations with $\lambdans>1000$ are
not affected significantly, we find that $\lambdans< 1000$ populations require 
$10-20\%$ more events to attain the same measurement accuracy as under
paradigm A. In either case, the accumulation of information from the general
NSBH population will likely be slower than from BNS inspirals~\cite{Mandel:2009pc,
Lackey2014,Wade:2014vqa,Agathos:2015a}, depending on the mass distribution of 
stellar-mass black holes. Though, template models for the latter may be more
uncertain due to missing point-particle PN terms at orders comparable to
tidal terms~\cite{Lackey2014}.
% 
We conclude that within as few as $20-30$ observations of disruptive NSBH
mergers, aLIGO will begin to place interesting bounds on NS deformability.
This, amongst other things, will allow us to rank different equations of 
state for neutron star matter from most to least likely, within a few years'
detector operation.
% 
Our methods and results are detailed in Sec.~\ref{s1:multiple_observations}.

@@

% ###########################################################
% ###########################################################
\section{Choice of underlying BBH model in LEA}
% ###########################################################
% 
\begin{figure}
\centering    
\includegraphics[width=1.05\columnwidth]{match-q-spin1-PhenomC-mf0_01.pdf}
\caption{%
We compare two alternatives of the tidal NSBH model from Ref.~\cite{
Lackey:2013axa}, which differ in their underlying BBH prescriptions. One which
we use in this study uses SEOBNRv2, while the other uses IMRPhenomC as its
base. In this figure, we show the normalized overlap (match) between the two for
$2,000,000$ points sampled uniformly in the NSBH parameter space. We find that
for $q\gtrsim 4$ the discrepancies between IMRPhenomC and SEOBNRv2 as reported
in~\cite{Kumar:2015tha} dominate over tidal terms.
}
\label{fig:PhenomC_vs_SEOBNRv2_LEA}
\end{figure}
%
The waveform model used in this paper is a variant of those calibrated
in Ref~\cite{Lackey:2013axa}. In that work, the authors also calibrate
a tidal prescription with the phenomenological model IMRPhenomC~\cite{
Santamaria:2010yb}
as the base BBH model. Previous work~\cite{Kumar:2015tha} has shown that 
IMRPhenomC can exhibit pathological behavior for mass-ratios ever so
$q\gtrsim 4$ and/or non-zero black hole spins. We compute noise-weighted
inner-products between the two variants for $2,000,000$ points sampled
over the NSBH parameter space, and show the results in Fig.~\ref{
fig:PhenomC_vs_SEOBNRv2_LEA}. We restrict the comparison to frequencies
that are affected by the tidal disruption of the NS, by integrating
the inner-products from $f = \mathrm{max}(15, 0.01/M)$~Hz
(where $M$ is expressed in seconds ($1M_\odot \simeq 4.925\mu$S, see
Eq.~(32-34) of~\cite{Lackey:2013axa}).
% 
We find that the differences between the two BBH approximants dominate
over the tidal terms that are added to their phasing. As SEOBNRv2 is
more reliable than IMRPhenomC, we recommend the use of SEOBNRv2-based
LEA+ in upcoming LIGO-Virgo analyses.

>>>>

\begin{abstract}
The pioneering discovery of gravitational waves (GW) by Advanced LIGO has ushered
us into an era of observational GW astrophysics. Compact binaries remain the 
primary target sources for GW observation, of which neutron star - black hole 
(NSBH) binaries form an important subset.
% 
GWs from NSBH sources carry signatures of (a) the tidal distortion of the
neutron star by its companion black hole during inspiral, and (b) its
potential tidal disruption near merger.
%
In this paper, we present a Bayesian study of the measurability of
neutron star tidal deformability $\lambdans\propto (R/M)_\mathrm{NS}^{5}$
using observation(s) of inspiral-merger GW signals from disruptive NSBH
coalescences, taking into account the crucial effect of black hole spins.
% 
First, we find that if non-tidal templates are used to estimate source 
parameters for an NSBH signal, the bias introduced in the estimation of
non-tidal physical parameters will only be significant for loud signals with
signal-to-noise ratios greater than $\simeq30$.
% 
For similarly loud signals, we also find that we can begin to put
interesting constraints on $\lambdans$ (factor of $1-2$) with individual
observations.
% 
Next, we study how a population of realistic NSBH detections will improve our
measurement of neutron star tidal deformability. For an astrophysically likely
population of {\it disruptive} NSBH coalescences, we find that $20-35$ events
are sufficient to constrain $\lambdans$ within $\pm 25-50\%$, depending on the
neutron star equation of state. For these calculations we assume that LIGO will
detect black holes with masses within the astrophysical {\it mass-gap}. In case
the mass-gap remains {\it preserved} in NSBHs detected by LIGO, we estimate that
approximately $25\%$ additional detections will furnish comparable $\lambdans$
measurement accuracy.
% 
In both cases, we find that it is the loudest $5-10$ events that provide
most of the tidal information, and not the combination of tens of low-SNR
events, thereby facilitating targeted numerical-GR follow-ups of NSBHs.
% 
We find these results encouraging, and recommend that an effort to 
measure $\lambdans$ be planned for upcoming NSBH observations with the
LIGO-Virgo instruments.
\end{abstract}

@@

The EoS of matter at nuclear densities is believed to be universal among all
neutron stars. The Tolman-Oppenheimer-Volkoff equation~\cite{Tolman:1939jz,
Oppenheimer:1939ne,1934PNAS...20..169T}
would then predict that NS properties satisfy a universal relationship
between $\lambdans$ and $\mns$. As the final part of
this paper, we combine information from multiple observations of realistic NSBH
systems and perform a fully-Bayesian analysis of how our estimation of
$\lambdans$ changes as we accumulate detections. This is similar to an earlier
study~\cite{DelPozzo:13} aimed at binary neutron stars.
% % 
We restrict ourselves to a population of NSs with masses clustered very tightly
around $1.35M_\odot$ (with a negligible variance), and negligible spins. We
sample different nuclear EoSs by sampling entire populations fixing different
values for the NS tidal deformability.
% These values correspond to a hard EoS at the lower end, and a soft EoS at the upper.
For all populations, we take source locations to be uniformly distributed in
spatial volume, and source orientations to be uniform on the $2-$sphere. To
summarize, we find the following:
% 
(a) Our median estimate for $\lambdans$ starts out prior dominated, but 
converges to within $10\%$ of the true value within $10-20$ detections.
(b) Measurement uncertainties for $\lambdans$, on the other hand, depend on
$\lambdans$ itself. We find that for hard equations of state (with 
$\lambdans\geq 1000$), $10-20$ observations are sufficient to constrain
$\lambdans$ within $\pm 50\%$. For softer equations of state, the same level
of certainty would require substantially more ($25-40$) observations.
% 
(c) Further, if the astrophysical ``mass-gap''~\cite{Bailyn:1997xt,
Kalogera:1996ci,Kreidberg:2012,Littenberg:2015tpa} is real, we find that $20-50\%$
additional observations would be required to attain the same measurement
accuracy as above. (d) Putting tighter constraints on the $\lambdans$ of a
population would require $50+$ NSBH observations, in any scenario.
And, (e) it is the loudest $5-10$ events that will furnish the bulk
of tidal information, and not the combination of a large number of 
low-SNR events.
% 
All of the above is possible within a few years of design
aLIGO operation~\cite{Abadie:2010cfa}.

@@

\includegraphics[width=1.025\columnwidth,trim=1cm 0 0 0]{LambdaMedianCIWidths90pc_vs_N_AllPopulation_SNRSorted}
\includegraphics[width=1.025\columnwidth,trim=0 0 1cm 0]{LambdaCIWidths90pc_vs_N_AstroPopulation_SNRSorted}
\caption{This figure is similar to Fig.~\ref{fig:TT_LambdaError_vs_N_L500_2000_CI90_0_AllInOne},
with the only difference being that events in each population
have been sorted according to their signal strength (SNR), instead of
their simulated chronology. We note that information about
the tidal deformability of neutron stars comes primarily 
from the loudest $5-10$ events, whether we allow BH masses
in the mass gap (left panel) or restrict them to
$\mbh\geq 5M_\odot$ (right panel).
\label{fig:TT_LambdaError_vs_N_L500_2000_CI90_0_AllInOne_SNRSorted}

@@

{\it Do all events matter?} In Ref.~\cite{Lackey2014} the authors demonstrate
that the overwhelming majority of information about the NS equation of state
comes from the loudest $\sim 5$ events in the case of binary neutron star
detections, and not from the combined effect of a large number of low-SNR
systems.
The question naturally arises if the same is true for NSBH sources as well.
Therefore, in Fig.~\ref{fig:TT_LambdaError_vs_N_L500_2000_CI90_0_AllInOne_SNRSorted}
we re-evaluate the accumulation of information with each successive NSBH 
detection, sorting the events in each population by their SNR instead of 
simulated chronology. We find the same qualitative behavior as in the case of 
BNSs~\cite{Lackey2014}. 
Whether or not there is an astrophysical mass gap, we find
that the bulk of tidal information will be furnished by the loudest $5-10$
NSBH detections of aLIGO detectors. This result is especially encouraging
to NR follow-up efforts for GW detections, as we now know that only a handful
of loudest NSBH events will merit full numerical-GR follow-up simulations.

To summarize, in this section we study the improvement in our measurement of 
NS deformability parameter $\lambdans$ with an increasing number of events. We
do so by simulating plausible populations of disrupting NSBH binaries (with
$\rho\geq 10$). We find that:
(i) for more deformable neutron stars (harder equation of states), the median
value of $\lambdans$ comes within $10\%$ of the true value with as 
few as $10$ events, while achieving the same accuracy for softer equations of 
state will take $15-20$ source detections; (ii) the statistical uncertainty
associated with $\lambdans$ measurement shrinks to within $\pm50\%$ with
$10-20$ events, and to within $\pm 25\%$ with $50+$ events, when source 
$\lambdans\geq 1000$; (iii) for softer equations of state, the same could take
$25-40$ and $50+$ events, respectively for the two uncertainty thresholds;
and (iv) if BHs really do observe the astrophysical mass-gap, the information
accumulation is somewhat slower than if they do not. We conclude that within
$20-30$ observations, aLIGO would begin to place very interesting bounds on 
the NS deformability, which would allow us to rule out or rank different
equations of state for neutron star matter. Within this population, we also
find that it will be the loudest $5-10$ events that will furnish most of the
tidal information. Our key findings are summarized in 
Fig.~\ref{fig:TT_LambdaMedian_vs_N_AllInOne} -
\ref{fig:TT_LambdaError_vs_N_L500_2000_CI90_0_AllInOne_SNRSorted}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}\label{s1:discussion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

We probe two astrophysical paradigms, one that allows for BH masses to lie
within the astrophysical mass-gap (paradigm A), and one that does not (paradigm
B).
% 
{\it For paradigm A}, we find the following: (i) for the softer equations of
state that result in less deformable neutron stars, $15-20$ detections bring
the measured probability distribution for $\lambdans$ entirely within the prior,
which ensures that the median $\lambdans$ tracks the true value to within $10\%$.
(ii) For NSBH populations with more deformable NSs ($\lambdans> 1000$),
the same is achievable within as few as $10$ (or $15$ at most) realistic
observations. (iii) The statistical uncertainty associated with $\lambdans$
measurement can be restricted to be within $\pm50\%$ using $10-20$ observations
when $\lambdans> 1000$), and using $25-40$ observations for softer equations
of state. All of the above is possible within a few years of design
aLIGO operation~\cite{Abadie:2010cfa}, if astrophysical BHs are allowed
masses $< 5M_\odot$ (i.e. in the mass-gap). However, further
restricting $\lambdans$ will require $50+$ NSBH observations.
% 
{\it For paradigm B}, we find the information accumulation to be somewhat slower.
While the quantitative inferences for populations with $\lambdans>1000$ are
not affected significantly, we find that $\lambdans< 1000$ populations require 
$10-20\%$ more events to attain the same measurement accuracy as under
paradigm A. In either case, the accumulation of information from the general
NSBH population will likely be slower than from BNS inspirals~\cite{Mandel:2009pc,
Lackey2014,Wade:2014vqa,Agathos:2015a}, depending on the mass distribution of 
stellar-mass black holes. Though, template models for the latter may be more
uncertain due to missing point-particle PN terms at orders comparable to
tidal terms~\cite{Lackey2014}.
% 
We conclude that within as few as $20-30$ observations of disruptive NSBH
mergers, aLIGO will begin to place interesting bounds on NS deformability.
This, amongst other things, will allow us to rank different equations of 
state for neutron star matter from most to least likely, within a few years'
detector operation.
% 
We also find that, within this population, the loudest $\sim 5-10$ events
will provide us with most of the tidal information, and will therefore
merit full NR follow-up.
% 
Our methods and results are detailed in Sec.~\ref{s1:multiple_observations}.

@@

% ###########################################################
% ###########################################################
\section{Choice of underlying BBH model in LEA}
% ###########################################################
% 
\begin{figure}
\centering    
\includegraphics[width=1.05\columnwidth]{match-q-spin1-PhenomC-mf0_01.pdf}
\caption{%
We compare two alternatives of the tidal NSBH model from Ref.~\cite{
Lackey:2013axa}, which differ in their underlying BBH prescriptions. One which
we use in this study uses SEOBNRv2, while the other uses IMRPhenomC as its
base. In this figure, we show the normalized overlap (match) between the two for
$2,000,000$ points sampled uniformly in the NSBH parameter space. We find that
for $q\gtrsim 4$ the discrepancies between IMRPhenomC and SEOBNRv2 as reported
in~\cite{Kumar:2015tha} dominate over tidal terms.
\label{fig:PhenomC_vs_SEOBNRv2_LEA}
}
\end{figure}
%
The waveform model used in this paper is a variant of those calibrated
in Ref~\cite{Lackey:2013axa}. In that work, the authors also calibrate
a tidal prescription with the phenomenological model IMRPhenomC~\cite{
Santamaria:2010yb}
as the base BBH model. Previous work~\cite{Kumar:2015tha} has shown that 
IMRPhenomC can exhibit pathological behavior for mass-ratios ever so
$q\gtrsim 4$ and/or non-zero black hole spins. We compute noise-weighted
inner-products between the two variants for $2,000,000$ points sampled
over the NSBH parameter space, and show the results in Fig.~\ref{fig:PhenomC_vs_SEOBNRv2_LEA}. We restrict the comparison to frequencies
that are affected by the tidal disruption of the NS, by integrating
the inner-products from $f = \mathrm{max}(15, 0.01/M)$~Hz
(where $M$ is expressed in seconds ($1M_\odot \simeq 4.925\mu$S, see
Eq.~(32-34) of~\cite{Lackey:2013axa}).
% 
We find that the differences between the two BBH approximants dominate
over the tidal terms that are added to their phasing. As SEOBNRv2 is
more reliable than IMRPhenomC, we recommend the use of SEOBNRv2-based
LEA+ in upcoming LIGO-Virgo analyses.
